---
title: "Project Title"
author: 
- Lenny Budlinger
- Thomas van der Wardt 
- Oliver Zingg

format:
  html:
    toc: true
    number_sections: true
    toc_float: true
    toc_depth: 5
    theme: cosmo
    highlight: zenburn
---

# Packages

```{r}
#| code-fold: true
#| code-summary: "Show the code"
#| warning: false
#| error: false

library(tidyverse) #ggplot for plotting, dplyr for data manipulation


```

# Dataset

Short description

## Load

```{r}

```

# Key take aways

## Problems with modelling and data

## Interpretations of results

# Modelling

## Linear Model

## **Short description of model:**

## **What have we learned?:**

## GLM Poisson

## **Short description of model:**

## **What have we learned?:**

## GLM Binomial

## **Short description of model:**

## **What have we learned?:**

## GAM

## **Short description of model:**

## **What have we learned?:**

## Neural Network

## **Short description of model:**

## **What have we learned?:**

## Support Vector Machine

## **Short description of model:**

## **What have we learned?:**

