---
title: "Project Title"
author: 
- Lenny Budlinger
- Thomas van der Wardt 
- Oliver Zingg

format:
  html:
    toc: true
    number_sections: true
    toc_float: true
    toc_depth: 5
    theme: cosmo
    highlight: zenburn
---

# Packages

```{r}
#| code-fold: true
#| code-summary: "Show the code"
#| warning: false
#| error: false

# Install Packages ---------------------------

# install.packages("tidyverse")
# install.packages("lubridate")
# install.packages("gt")
# Load packages ---------------------------

library(tidyverse) # ggplot for plotting, dplyr for data manipulation
library(lubridate) # package for handeling time variable
library(gt) # Print tables

```

# Dataset

Short description

## Load

```{r}
#| echo: false 
#| code-fold: true
#| code-summary: "Show the code"
#| warning: false

# Load data ---------------------------

# get working directory if csv not loading or use absolute path
# getwd() 

df_accidents <- read.csv("./../Data/global_traffic_accidents.csv")

df_accidents %>% 
  head() %>% 
  gt(rownames_to_stub = TRUE)
```

## Transform Data

```{r}
# Time variable ---------------------------
# Split day, month, year

df_accidents <- df_accidents %>% 
  # creates new columns for day, month, year 
  mutate(
    Day = day(Date),
    Month = month(Date),
    Year = year(Date)
  ) %>%
  # moves new created columns after Date
  relocate(Day, Month, Year, .after = Date)
  
# Location variable ---------------------------
# City, Country

df_accidents <- df_accidents %>% 
  # separate by ","
  separate(Location, into = c("City", "Country"), sep = ",")

# First 5 rows ---------------------------
df_accidents %>% 
  head() %>% 
  gt(rownames_to_stub = TRUE)
```

```{r}

```

## Explore data

### Impact of seasons?
```{r}
df_accidents %>% 
  group_by(Year, Month) %>%
  summarise(sum_Casualties = sum(Casualties), .groups = "drop") %>% 
  ggplot(
    aes(x = Month, y = sum_Casualties)
  ) + 
  geom_line() + 
  scale_x_continuous(limits = c(1, 12), breaks = seq(1, 12, by = 1)) +  # Custom ticks
  theme_minimal()  +
  facet_wrap(~Year) 
```


# Key take aways

## Problems with modelling and data

## Interpretations of results

# Modelling

## Linear Model

**Short description of model:**

**What have we learned?:**

## GLM Poisson

**Short description of model:**

**What have we learned?:**

## GLM Binomial

**Short description of model:**

**What have we learned?:**

## GAM

**Short description of model:**

**What have we learned?:**

## Neural Network

**Short description of model:**

**What have we learned?:**

## Support Vector Machine

**Short description of model:**

**What have we learned?:**

